---
title: "STA 9750 - QIAN LIN"
---

<div class="intro-center">
  <h1>Welcome to My STA 9750 Project Portfolio</h1>
  <p>Hello — I’m <b>QIAN LIN</b>, a Master’s student at Baruch College.</p>
  <a class="btn btn-primary" href="files/QianLin_Resume.pdf" target="_blank">📄 Download My Resume</a>
</div>

# Dynamic code / Last updated

```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`

```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```


# Mini Projects

<div class="project-card">
  <h3>Mini Project 1 — Netflix Top 10 Analysis</h3>
  <p>Exploring streaming trends using title-level analytics</p>
  <a href="mp01.html">View Project →</a>
</div>

<div class="project-card">
  <h3>Mini Project 2 — Housing, Income & Growth</h3>
  <p>ACS socioeconomic trends and NYC building-permit patterns</p>
  <a href="mp02.html">View Project →</a>
</div>

<div class="project-card">
  <h3>Mini Project 3 — NYC Trees & Council Districts</h3>
  <p>Spatial analysis linking tree density and district-level characteristics</p>
  <a href="mp03.html">View Project →</a>
</div>

<div class="project-card">
  <h3>Mini Project 4 — CES Employment & Revisions</h3>
  <p>Time series trends and BLS revision patterns</p>
  <a href="mp04.html">View Project →</a>
</div>

# Individual Project Report

<div class="project-card">
  <h3>Traffic Volume, Road Type, and Street Damage in New York City</h3>
  <p>
    An individual analytical report examining how traffic volume and road
    characteristics are associated with street damage complaints across NYC.
  </p>
  <a href="Individual_Report.html">View Report →</a>
</div>

# Group Project Report

<div class="project-card">
  <h3>Traffic, Complaints, and Urban Infrastructure in New York City</h3>
  <p>
    A collaborative group analysis examining relationships between traffic
    volume, noise, air quality, and street-condition complaints across NYC.
  </p>
  <a href="https://xiuyinglan.github.io/STA9750-2025-FALL/group_report.html" target="_blank">
    View Group Report →
  </a>
</div>

<div class="footer">
  ✨ Thank you for visiting my STA 9750 project portfolio! ✨  
  <br><br>
  © 2025 Qian Lin — Baruch College
</div>




